Seeing Double: Transient Dips and Photometric Binarity in PTFO$\,$8-8695

### 2020-04-20 paper/ms.tex
update paper

%\title{Against the Planetary Interpretation of PTFO$\,$8-8695b}
% \title{Seeing Double: Still Against the Planetary Interpretation of
% PTFO$\,$8-8695b}
% \title{Seeing Double: TESS and Gaia Show That PTFO$\,$8-8695b is Unlikely a
% Planet}
%\title{Seeing Double: Indications From {\it TESS} and {\it Gaia} That
%PTFO$\,$8-8695b is Not a Planet}
\title{Seeing Double: PTFO$\,$8-8695 is a Photometric Binary With One
Component Showing Transient Dips }

@@

\begin{abstract}
  PTFO$\,$8-8695b is a candidate hot Jupiter in the 7--10 million year old
  Orion-OB1a cluster. We inspected data from TESS and Gaia to clarify whether
  it is actually a planet.  The TESS lightcurve shows that the dominant
  variability in this system is a sinusoid with a ``long'' period $P_{\rm
  \ell}=11.96\,$hr, likely caused by stellar rotation.  Also present is a
  complex signal, previously identified as the planet candidate, that repeats
  with a ``short'' period $P_{\rm s}= 10.74\,$hr.  The two signals beat every
  4.48 days.  Although there is a dip in the short-period signal, ground-based
  photometry from the past decade shows that the orbital phase of the dip seems
  to have instantaneously jumped, at least once, and perhaps twice.  Planets do
  not ``jump'' in orbital phase.  Furthermore, the Gaia data show that
  PTFO$\,$8-8695 is probably a photometric binary.  Given the evidence, we
  believe that PTFO$\,$8-8695 is a binary M dwarf in which one star shows the
  ``long'' rotation signal, and the other is showing ``transient dipping'' that
  has also been observed in other young M dwarfs.  The planetary interpretation
  seems unlikely.
\end{abstract}

%TODO
\keywords{
	Exoplanet evolution (491),
	Open star clusters (1160),
	Stellar ages (1581),
	Stellar rotation (1629),
	Variable stars (1761)
}

@@

Another first could be the direct detection of H$\alpha$ emission from
the planet itself \citep{johnskrull_h_2016}.  While the stellar
chromosphere emits in H$\alpha$, it seems that there is an additional
excess H$\alpha$ emission that could be in phase with the planetary
orbit.  The average velocity width of the excess H$\alpha$ emission is
87$\,$km$\,$s$^{-1}$, and its equivalent width is 70-80\% that of the
stellar chromosphere \citep{johnskrull_h_2016}.  The proposed
explanation is that the outflowing mass from the planet may explain
this excess emission as well \citep{johnskrull_h_2016}.

There are perhaps a few challenges to the planetary interpretation (if
these ``features'' are not already seen as such).  They include that
the planet does not seem to emit infrared radiation in occulation, at
least anywhere near the expected amplitude \citep{yu_tests_2015}.  In
addition, despite measurement attempts by multiple investigators, it
does not seem to show the Rossiter effect at the amplitude expected
given the rapid stellar rotation and large planet size
\citep{yu_tests_2015,ciardi_followup_2015}.  Finally, detailed
modelling of the ``precession + gravity darkening'' transits has shown
that the necessary degree of gravity darkening is too great, given the
spectroscopically observed equatorial velocity
\citep{howarth_reappraisal_2016}.  Additionally, as the
gravity-darkened star precessed about its rotation axis, it would need
to show photometric variability that has not been observed.

While the planetary interpetation has challenges, alternative
explanations do as well.  High-latitude accretion hotspots might
produce the observed H$\alpha$ variability, but \ptfo\ does not have
an infrared (IR) excess associated with the presence of an inner disk.
Low-latitude starspots, hot or cold, struggle to produce the necessary
dip durations.  A circumstellar cloud of dust or gas might be
entrained in the stellar magnetosphere, but clumps would need to be
invoked to explain the short-duration flux dips.

In the following, Section~\ref{sec:observations} presents newly
available observations from the Transiting Exoplanet Survey Satellite
(TESS; \citealt{ricker_transiting_2015}) and Gaia
\citep{gaia_collaboration_gaia_2018}.  Analyzing the TESS data in
Section~\ref{sec:tess}, we find two distinct signals.  A long sinusoid
repeats every 12.0 hours, and is probably stellar rotation.  A short
``dip + complex modulation'' repeats every 10.7 hours.  Analyzing the
Gaia data in Section~\ref{sec:gaia}, we show that \ptfo\ is quite
likely a photometric binary.  In Section~\ref{sec:discussion}, we note
that the orbital phase of the dip seems to have instantaneously
changed, perhaps twice.  We argue that PTFO$\,$8-8695 is a binary M
dwarf in which one star shows a rotation signal, and the other is
showing ``transient dipping'' that has been observed in other young M
dwarfs, and is likely caused by eclipses of warm particulate clouds.
Section~\ref{sec:conclusions} summarizes our main points.

@@

%\begin{figure*}[t!]
%	\begin{center}
%		\leavevmode
%		\includegraphics[width=0.9\textwidth]{f6_comp.png}
%	\end{center}
%	\vspace{-0.7cm}
%	\caption{ {\bf foo.}
%	bar
%		\label{fig:corner}
%	}
%\end{figure*}

@@

Between July 25, 2014 and May 23, 2016, Gaia acquired 121 ``good''
observations of \ptfo, {\it i.e.}, observations that were not strongly
downweighted in the astrometric solution of the source (CITE).
%TODO: give an original explanation of how Gaia works
The Gaia data processing team at XXXINSTITUTION combined the XXX and
YYY into a global astrometric solution, from which the parallax of 
\ptfo\ was derived (CITE).
The star was assigned a Gaia DR2 identifier of 3222255959210123904
(CITE).
In addition, measurements of the stellar proper motion between
observation epochs were made.
%FIXME sentence is bad

>>>>

%\title{Against the Planetary Interpretation of PTFO$\,$8-8695b}
% \title{Seeing Double: Still Against the Planetary Interpretation of
% PTFO$\,$8-8695b}
% \title{Seeing Double: TESS and Gaia Show That PTFO$\,$8-8695b is Unlikely a
% Planet}
%\title{Seeing Double: Indications From {\it TESS} and {\it Gaia} That
%PTFO$\,$8-8695b is Not a Planet}
\title{Seeing Double: Transient Dips and Photometric Binarity in
PTFO$\,$8-8695}

@@

\begin{abstract}
  PTFO$\,$8-8695b is a candidate hot Jupiter in the 7--10 million year
  old Orion-OB1a cluster. We inspected data from TESS and Gaia to
  clarify whether it is actually a planet.  The Gaia data show that
  PTFO$\,$8-8695 is a photometric binary with respect to members of
  its kinematic group.  The TESS lightcurve shows that the dominant
  variability in this system is a sinusoid with a ``long'' period
  $P_{\rm \ell}=11.96\,$hr, presumably caused by stellar rotation.
  Also present is a complex signal, previously identified as the
  planet candidate, that repeats with a ``short'' period $P_{\rm s}=
  10.74\,$hr.  The two signals beat every 4.48 days.  Although there
  is a dip in the short-period signal, ground-based photometry from
  the past decade shows that the orbital phase of the dip seems to
  have instantaneously jumped, at least once, and perhaps twice.
  Planets do not ``jump'' in orbital phase.  Given the evidence, we
  believe that PTFO$\,$8-8695 is a binary M dwarf in which one star
  shows the long rotation signal, and the other star is showing
  ``transient dipping'' that has been seen in a few other young
  weak-lined T Tauri stars.  The origin of these transient dips is
  still undetermined, but our preferred explanation is eclipses of
  clouds of gas or dust at the Keplerian co-rotation radius.
\end{abstract}

\object{PTFO 8-8695}
\object{PTFO 8-8695b}
\object{CVSO 30}

\keywords{
	Exoplanet evolution (491),
  Pre-main sequence stars (1290),
	Stellar ages (1581),
	Stellar rotation (1629),
	Variable stars (1761),
  Low mass stars (2050)
}

@@

Another first could be the direct detection of H$\alpha$ emission from
the planet itself \citep{johnskrull_h_2016}.  While the stellar
chromosphere emits in H$\alpha$, it seems that there is an additional
excess H$\alpha$ emission that could be in phase with the planetary
orbit.  The average velocity width of the excess H$\alpha$ emission is
87$\,$km$\,$s$^{-1}$, and its equivalent width is 70-80\% that of the
stellar chromosphere \citep{johnskrull_h_2016}.  The proposed
explanation is that outflowing mass from the planet may explain this
excess emission as well \citep{johnskrull_h_2016}.

There are perhaps a few challenges to the planetary interpretation (if
these ``features'' are not already seen as such).  They include that
the planet does not seem to emit infrared radiation in occulation, at
least anywhere near the expected amplitude \citep{yu_tests_2015}.  In
addition, despite measurement attempts by multiple investigators, it
does not seem to show the Rossiter effect at the amplitude expected
given the rapid stellar rotation and large planet size
\citep{yu_tests_2015,ciardi_followup_2015}.  Finally, detailed
modelling of the ``precession + gravity darkening'' transits has shown
that the necessary degree of gravity darkening is too great, given the
spectroscopically observed equatorial velocity
\citep{howarth_reappraisal_2016}.  Additionally, as the
gravity-darkened star precessed about its rotation axis, it would need
to show photometric variability that has not been observed.  While the
planetary interpetation clearly faces challenges, alternative
explanations do as well.  High-latitude accretion hotspots might
produce the observed H$\alpha$ variability, and require fine-tuning
produce dips of the approach duration. Furthermore, \ptfo\ does not
have an infrared (IR) excess associated with the presence of an inner
disk \citep[{\it e.g.},][Figure~18]{yu_tests_2015}.  Low-latitude
starspots, hot or cold, struggle to produce the necessary dip
durations.

Alternatively, between 0.1\% and 1\% of rapidly rotating low-mass
stars in $\mathcal{O}$(10)$\,$Myr old associations show narrow dips in
phase with strong stellar rotation signals \citep{rebull_usco_2018}.
The dips can persist over months, but their depths often vary, and
sometimes change immediately after stellar flares.  The explanation
proposed by {\it e.g.}, \citet{stauffer_orbiting_2017} and
\citet{david_transient_2017} is that a circumstellar cloud of dust or
gas might be orbiting near the Keplerian co-rotation radius.  For this
explanation to be viable for \ptfo, a clear determination of the
stellar rotation period is necessary.  To date, this has not been
possible
\citep{van_eyken_ptf_2012,koen_multicolour_2015,raetz_yeti_2016}.

We begin in Section~\ref{sec:observations} by describing newly
available observations from the Transiting Exoplanet Survey Satellite
(TESS; \citealt{ricker_transiting_2015}) and Gaia
\citep{gaia_collaboration_gaia_2018}.  The TESS lightcurve shows two
distinct signals, which we extract and analyze in
Section~\ref{sec:tess}.  A long-period sinusoid repeats every 12.0
hours, and is probably stellar rotation.  A short-period ``dip +
complex modulation'' repeats every 10.7 hours.  Analyzing the Gaia
data in Section~\ref{sec:gaia}, we show that relative to its kinematic
group, \ptfo\ is a photometric binary.  We collect and discuss the
puzzle pieces in Section~\ref{sec:discussion}.  The orbital phase of
the dip seems to have instantaneously changed over the past decade.
In addition, a number of other young stars show similar lightcurve
morphologies.  We therefore argue that PTFO$\,$8-8695 is a binary M
dwarf in which one star shows a rotation signal, and the other is
showing ``transient dipping'' that is likely caused by eclipses of
clouds of material at the Keplerian co-rotation radius.
Section~\ref{sec:conclusions} summarizes our main points.

@@

@@

Between July 25, 2014 and May 23, 2016, Gaia measured about 300
billion centroid positions of 1{.}6 billion stars
\citep{gaia_collaboration_gaia_2016,lindegren_gaiasoln_2018,gaia_collaboration_gaia_2018}.
These CCD observations were used to estimate positions, proper
motions, and parallax for the brighest 1{.}3 billion stars, including
\ptfo.  There were 121 ``good'' observations of \ptfo, {\it i.e.},
observations that were not strongly downweighted in the astrometric
solution of the source.  The star was assigned a Gaia DR2 identifier
of 3222255959210123904.  It photometric brightness was measured using
selected bands ($G$, $Rp$, and $Bp$) of the Gaia Radial Velocity
Spectrometer \citep{cropper_gaia_2018,evans_gaia_2018}.

%FIXME FROM HERE
